Nitric oxide feedback to ciliary photoreceptor cells gates a UV avoidance circuit

### 2023-06-13 Manuscript/Manuscript.Rmd
MS: added sulfolink protocol and link to protocol on bio-protocols
MS: modelling part edits

Nitric oxide (NO) is one such modulator with unique physical and signalling properties.
This free radical synthesized from L-arginine by nitric oxide synthase (NOS) is short-lived and can diffuse across biological membranes [@Cudeiro1999; @thomas2015].
Canonical NO signalling involves the Ca2+/calmodulin-dependent activation of NOS, NO production and diffusion, and the NO-dependent activation of soluble guanylate cyclases (sGC).
This can occur cell-autonomously or in other cells leading to cGMP production [@Bredt1990; @Hoelscher1997].
Given that NOS activation is calcium dependent and NOS shows neuron-type-specific expression [@Gibbs1998a; @Wildemann1999; @Aso2019; @Mobley2022], NO action can lead to the activity-dependent modulation of neural circuits at specific sites [@Wang2007; @Vielma2014; @jacoby2018; @Aso2019] .

@@

To further analyse the dynamics of responses to UV light and formally describe cPRC phototransduction, synaptic connections and NO retrograde signalling, we developed a mixed cellular-circuit-level mathematical model.
We used our Ca2+ imagining data of cPRC, INNOS and INRGWa cells collected in wild type, *NOS* knockout, and *NIT-GC2* morphant larvae to formulate assumptions that are the basis of the proposed model.

We model a direct c-opsin1-dependent response to UV leading to the initial rise in Ca2+ levels.
The mechanism of this signal is not known but may include Gα and Gβ signalling, CNGAα channel activation or other pathways.
We further assume that in cPRC cells, Ca2+ levels increase with cGMP.
We model a decrease in cGMP and Ca2+ levels due to a NIT-GC2-dependent pathway (Figre 6A).
The effect of NO in cPRC cells is captured by another term that describes a NIT-GC1 and NO-dependent increase in cGMP.
Based on the synaptic connectome, we infer a feedforward coupling between cPRC cells and INNOS and INRGWa cells.
The proposed synaptic coupling is UV-dependent, decays linearly and is suppressed in NIT-GC2 morphants (Figure 6A).

In the INNOS cells, we assume that excitatory synaptic input from cPRC (or rebound from tonic inhibition) leads to a rise in Ca2+ leading to NO production via NOS.
This input is NIT-GC2-dependent.
In the INRGWa cells, we model a decrease in Ca2+ based on an inhibitory UV-dependent synaptic signal from the cPRCs.
We further assume the existence of direct excitatory coupling between cPRC Ca2+ levels and INRGWa Ca2+ to account for the late effects of cPRC on INRGWa (Figure 6A).
The UV-dependent and Ca2+-dependent signals may be mediated by different transmitters released by cPRCs during distinct phases of their activation cycle.
In addition, we assume a direct inhibitory coupling between INNOS Ca2+ and INRGWa Ca2+.
Finally, in all variables, we assumed a linear decay and constant production to set a steady state (Figure 6A).
Since the aim of the model is to capture the normalised fluorescence data, the model is nondimensionalised.
UV stimulation is modelled as a square pulse.
To find model parameters producing an output fitting the experimental data we employed a global optimisation method known as a genetic algorithm (GA).

Through this optimisation procedure, we found parameters that gave a very good fit to our experimental data under all conditions.
We could fit to mean Ca2+ values but also individual calcium transients (Figure 6B, Figure 6--figure supplement 1).
Our model only includes a minimal set of parameters and assumptions of interactions that are required to describe the dynamics of cPRC, INNOS and INRGWa in wild type and loss-of-function conditions (Figure 6A, B).
The model highlights that cPRC phototransduction employs distinct pathways that operate on different time scales and differentially influence cPRC Ca2+ levels.
The coupling between cPRCs and interneurons also requires different signalling mechanisms, either through distinct neurotransmitters or receptors expressed in the different cells.

@@

To generate new insights from the model, we asked how the duration and intensity of UV stimulation influences signalling output, measured as the magnitude, duration and time to maximum of the NO-dependent Ca2+ signal in the cPRCs.
The model allowed us to test a large number of conditions that would not have been feasible in single-larva experiments.
We found that the onset, magnitude and duration of the NO-dependent signal scales with stimulus intensity and duration (Figure 6--figure supplement 2).
This suggests an integratory function of NO signalling that tunes circuit output to the strength of UV exposure.

@@

Based on our data herein and previous work we can now propose a more detailed model of cPRC phototransduction and neurotransmission.
The cPRCs have high basal Ca2+ and respond to UV/violet light dependent on c-opsin1 [@Veraszto2018].
c-opsin1 forms a bistable photopigment and signals through Gi/oα and Gβγ [@tsukamoto2017; @Veedin2021; @tsukamoto2023].
In heterologous systems, the Gβγ subunits released following c-opsin1 activation open GIRK channels inducing K+ efflux (hyperpolarisation) [@tsukamoto2017; @tsukamoto2023] and close voltage-gated Ca2+ channels, thereby reducing intracellular calcium levels [@tsukamoto2023].
A GIRK channel is also expressed in the cPRCs (Figure 6A).
The pathway for the first rapid cPRC activation phase following UV/violet stimulus is not known but may involve the activation of a CNGα channel expressed in the cPRCs [@Tosches2014].
For the second inhibitory phase of phototransduction, we identified a key requirement for ciliary-localised NIT-GC2.
The mechanisms may involve c-opsin1-dependent inhibition of tonic NIT-GC2 activity and the reduction of ciliary cGMP.

NIT-GC2-dependent signalling is required for the feedforward activation of the INNOS cells through an unknown transmitter.
The activation of INNOS leads to NOS activation and NO release, potentially through a canonical Ca2+-calmodulin pathway.
Our model predicts feedforward inhibiton from INNOS to INRGWa, possibly mediated by glycin transmission.
NO released by INNOS neurites activates NIT-GC1 in cPRC projections that could lead to cGMP production and the opening of CNGAα.
This late-phase cPRC activation is not directly dependent on the UV/violet signal and can happen post-stimulus.
The late-phase cPRC activation leads to INRGWa activation via an unknown transmitter that is likely different from the one acting on INNOS.

@@

Pressure-induced Ca2+ transients in cPRCs lack an inhibitory phase and late activation and resemble UV responses in NIT-GC2 morphants [@bezares-calderon2023].
This indicates that the differentiation of sensory cues by the multisensory cPRCs occurs already at the level of sensory signal transduction.
The different signalling pathways then likely result in the differential release of transmitters and modulators that are decoded by the postsynaptic interneuron circuit.
The complex transmitter phenotype of cPRCs could underlie such differential signal processing.

@@

```{r echo=FALSE}
library(readxl)
key_resource_tale <- read_excel("key_resource_tale/key_resource_tale.xls")
df_key_resource_tale <- as.data.frame(key_resource_tale)
knitr::kable(df_key_resource_tale, format = "html", booktabs = TRUE)
```

@@

Antibodies were raised in rats or rabbits against synthetic peptides containing an N-term Cys residue (Altabioscience). The same Cys-containing synthetic peptides were used for affinity purification.
Antibodies were affinity purified from sera on a SulfoLink Coupling Resin (Thermo Fischer) as previously described [@Conzelmann2012].

@@

@@

This work was funded by the Wellcome Trust (214337/Z/18/Z).
This project has received funding from the European Research Council (ERC) under the European Union's Horizon 2020 research and innovation programme (grant agreement No 101020792).
KJ has been supported by a JSPS Postdoctoral fellowhip, LAYG by a BBSRC Discovery fellowship (BB/W010305/1).

>>>>

Nitric oxide (NO) is one such modulator with unique physical and signalling properties.
This free radical synthesized from L-arginine by nitric oxide synthase (NOS) is short-lived and can diffuse across biological membranes [@Cudeiro1999; @thomas2015].
Canonical NO signalling involves the Ca^2+^/calmodulin-dependent activation of NOS, NO production and diffusion, and the NO-dependent activation of soluble guanylate cyclases (sGC).
This can occur cell-autonomously or in other cells leading to cGMP production [@Bredt1990; @Hoelscher1997].
Given that NOS activation is calcium dependent and NOS shows neuron-type-specific expression [@Gibbs1998a; @Wildemann1999; @Aso2019; @Mobley2022], NO action can lead to the activity-dependent modulation of neural circuits at specific sites [@Wang2007; @Vielma2014; @jacoby2018; @Aso2019] .

@@

To further analyse the dynamics of responses to UV light and formally describe cPRC phototransduction, synaptic connections and NO retrograde signalling, we developed a mixed cellular-circuit-level mathematical model.
We used our Ca^2+^ imagining data of cPRC, INNOS and INRGWa cells collected in wild type, *NOS* knockout, and *NIT-GC2* morphant larvae to formulate assumptions that are the basis of the proposed model.

We model a direct c-opsin1-dependent response to UV leading to the initial rise in Ca^2+^ levels.
The mechanism of this signal is not known but may include Gα and Gβ signalling, CNGAα channel activation or other pathways.
We further assume that in cPRC cells, Ca^2+^ levels increase with cGMP.
We model a decrease in cGMP and Ca^2+^ levels due to a NIT-GC2-dependent pathway (Figre 6A).
The effect of NO in cPRC cells is captured by another term that describes a NIT-GC1 and NO-dependent increase in cGMP.
Based on the synaptic connectome, we infer a feedforward coupling between cPRC cells and INNOS and INRGWa cells.
The proposed synaptic coupling is UV-dependent, decays linearly and is suppressed in NIT-GC2 morphants (Figure 6A).

In the INNOS cells, we assume that excitatory synaptic input from cPRC (or rebound from tonic inhibition) leads to a rise in Ca^2+^ leading to NO production via NOS.
This input is NIT-GC2-dependent.
In the INRGWa cells, we model a decrease in Ca^2+^ based on an inhibitory UV-dependent synaptic signal from the cPRCs.
We further assume the existence of direct excitatory coupling between cPRC Ca^2+^ levels and INRGWa Ca^2+^ to account for the late effects of cPRC on INRGWa (Figure 6A).
The UV-dependent and Ca^2+^-dependent signals may be mediated by different transmitters released by cPRCs during distinct phases of their activation cycle.
In addition, we assume a direct inhibitory coupling between INNOS Ca^2+^ and INRGWa Ca^2+^.
Finally, in all variables, we assumed a linear decay and constant production to set a steady state (Figure 6A).
Since the aim of the model is to capture the normalised fluorescence data, the model is nondimensionalised.
UV stimulation is modelled as a square pulse.
To find model parameters producing an output fitting the experimental data we employed a global optimisation method known as a genetic algorithm (GA).

Through this optimisation procedure, we found parameters that gave a very good fit to our experimental data under all conditions.
We could fit to mean Ca^2+^ values but also individual calcium transients (Figure 6B, Figure 6--figure supplement 1).
Our model only includes a minimal set of parameters and assumptions of interactions that are required to describe the dynamics of cPRC, INNOS and INRGWa in wild type and loss-of-function conditions (Figure 6A, B).
The model highlights that cPRC phototransduction employs distinct pathways that operate on different time scales and differentially influence cPRC Ca^2+^ levels.
The coupling between cPRCs and interneurons also requires different signalling mechanisms, either through distinct neurotransmitters or receptors expressed in the different cells.

@@

To generate new insights from the model, we asked how the duration and intensity of UV stimulation influences signalling output, measured as the magnitude, duration and time to maximum of the NO-dependent Ca^2+^ signal in the cPRCs.
The model allowed us to test a large number of conditions that would not have been feasible in single-larva experiments.
We found that the onset, magnitude and duration of the NO-dependent signal scales with stimulus intensity and duration (Figure 6--figure supplement 2).
This suggests an integratory function of NO signalling that tunes circuit output to the strength of UV exposure.

@@

Based on our data herein and previous work we can now propose a more detailed model of cPRC phototransduction and neurotransmission.
The cPRCs have high basal Ca^2+^ and respond to UV/violet light dependent on c-opsin1 [@Veraszto2018].
c-opsin1 forms a bistable photopigment and signals through Gi/oα and Gβγ [@tsukamoto2017; @Veedin2021; @tsukamoto2023].
In heterologous systems, the Gβγ subunits released following c-opsin1 activation open GIRK channels inducing K+ efflux (hyperpolarisation) [@tsukamoto2017; @tsukamoto2023] and close voltage-gated Ca^2+^ channels, thereby reducing intracellular calcium levels [@tsukamoto2023].
A GIRK channel is also expressed in the cPRCs (Figure 6A).
The pathway for the first rapid cPRC activation phase following UV/violet stimulus is not known but may involve the activation of a CNGα channel expressed in the cPRCs [@Tosches2014].
For the second inhibitory phase of phototransduction, we identified a key requirement for ciliary-localised NIT-GC2.
The mechanisms may involve c-opsin1-dependent inhibition of tonic NIT-GC2 activity and the reduction of ciliary cGMP.

NIT-GC2-dependent signalling is required for the feedforward activation of the INNOS cells through an unknown transmitter.
The activation of INNOS leads to NOS activation and NO release, potentially through a canonical Ca^2+^-calmodulin pathway.
Our model predicts feedforward inhibiton from INNOS to INRGWa, possibly mediated by glycin transmission.
NO released by INNOS neurites activates NIT-GC1 in cPRC projections that could lead to cGMP production and the opening of CNGAα.
This late-phase cPRC activation is not directly dependent on the UV/violet signal and can happen post-stimulus.
The late-phase cPRC activation leads to INRGWa activation via an unknown transmitter that is likely different from the one acting on INNOS.

@@

Pressure-induced Ca^2+^ transients in cPRCs lack an inhibitory phase and late activation and resemble UV responses in NIT-GC2 morphants [@bezares-calderon2023].
This indicates that the differentiation of sensory cues by the multisensory cPRCs occurs already at the level of sensory signal transduction.
The different signalling pathways then likely result in the differential release of transmitters and modulators that are decoded by the postsynaptic interneuron circuit.
The complex transmitter phenotype of cPRCs could underlie such differential signal processing.

@@

```{r echo=FALSE}
library(readxl)
key_resource_table <- read_excel("key_resource_table/key_resource_tale.xls")
df_key_resource_table <- as.data.frame(key_resource_table)
knitr::kable(df_key_resource_table, format = "html", booktabs = TRUE)
```

@@

Antibodies were raised in rats or rabbits against synthetic peptides containing an N-term Cys residue (Altabioscience).
The same Cys-containing synthetic peptides were used for affinity purification.
Antibodies were affinity purified from sera on a SulfoLink Coupling Resin (Thermo Fischer) as previously described [@Conzelmann2012].
View a detailed protocol here: <https://bio-protocol.org/exchange/preprintdetail?id=2333&type=3>

@@

## Mathematical modelling

The mathematical model for the wild type data comprises six dynamical equations:

\begin{align*}
\frac{d C_P(t)}{dt} &= 1 - \delta^{C_P} C_P(t) + K_{UV}^{C_P} UV(t) C_O(t)  + K_{G}^{C_P} G(t) , \\
\frac{dC_O(t)}{dt} &= 1 - \delta^{C_O} C_O(t) - K_{UV}^{C_O} UV(t) C_O(t), \\
\frac{dG(t)}{dt} &= 1 - \delta^{G} G(t) + K_{GC2}^G \left(1 - UV(t)\right) G(t) + K_{GC1}^G N(t) , \\
\frac{dS(t)}{dt} &= UV(t) - \frac{S(t)}{\tau_S}, \\
\frac{dC_N(t)}{dt} &= 1 - \delta^{C_N} C_N(t) + K_{S}^{C_N} S(t), \\
\frac{dN(t)}{dt} &= 1 - \delta^{N} N(t) + K_{C_N}^{N} C_N(t), \\
\frac{dC_R(t)}{dt} &= 1 - \delta^{C_R} C_R(t) - K_{S}^{C_R} S(t) + K_{C_P}^{C_R} C_P(t) - K_{C_N}^{C_R} C_N(t) C_R(t), \\
UV(t) &=
    \begin{cases}
        a, & t_{UV\: start} \leq t \leq t_{UV\: end}\\
        0, & \text{otherwise}
    \end{cases}   .
\end{align*}

@@

This work was funded by the Wellcome Trust (214337/Z/18/Z).
This project has received funding from the European Research Council (ERC) under the European Union's Horizon 2020 research and innovation programme (grant agreement No 101020792).
KJ has been supported by a JSPS Postdoctoral fellowhip, LAYG by a BBSRC Discovery fellowship (BB/W010305/1), PS by a Wellcome Trust Institutional Strategic Support Award (204909/Z/16/Z).
